---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(mediaplanr)
```

This walks the whole path: get a plan in, change it, fork scenarios, compare
them. It sticks to *what to type*. For *why the package is shaped this way* —
what a line item is, why a plan cannot tell you what is historical — read
`vignette("plan_concepts")`.

## A plan in three lines

The smallest useful plan is a table with something to key on and a spend column.

```{r}
simple <- media_plan_from_df(
  data.frame(
    channel       = c("TV", "Search", "Social"),
    planned_spend = c(80000, 40000, 40000)
  ),
  grain = "channel",
  name  = "Q3 sketch"
)

simple
```

`name` is required — every plan carries one. Everything else is optional.

## A realistic plan

Real plans are weekly and keyed by more than channel. Here is a quarter at
`channel + partner + week`:

```{r}
weeks <- seq(as.Date("2026-04-06"), by = "week", length.out = 6)

plan_df <- expand.grid(
  week    = weeks,
  partner = c("NBC", "ESPN", "Google", "Meta"),
  stringsAsFactors = FALSE
)
plan_df$channel <- c(NBC = "TV", ESPN = "TV",
                     Google = "Search", Meta = "Social")[plan_df$partner]
plan_df$planned_spend <- c(42000, 28000, 19000, 23000)[
  match(plan_df$partner, c("NBC", "ESPN", "Google", "Meta"))
]

base <- media_plan_from_df(
  plan_df,
  grain      = c("channel", "partner", "week"),
  week       = "week",
  name       = "Q2 2026 Brand Plan",
  nickname   = "baseline",
  advertiser = "Acme Corp",
  planner    = "R. Roe",
  status     = "approved"
)

base
```

Two things to notice in that call:

- `grain` names the columns that identify a **row**. Order it coarsest-first;
  `roll_up()` later relies on that nesting.
- `week` marks which grain column is the week. It is validated as a `Date`, so
  ordering and range checks work without re-parsing.

`@data` stays an ordinary data frame throughout. The class adds guarantees, not
a wrapper you have to fight:

```{r}
head(base@data, 3)
sum(base@data$planned_spend)
```

## Editing: three shapes, one door

`build_scenario()` is the only way to change spend, and it always returns a
**new** plan. `edits` accepts three shapes; they exist because three different
callers need different things.

### 1. Operations — for a UI or an LLM

State what you want done and let R compute it. `target` may name any **subset**
of the grain, so one operation reaches many rows.

```{r}
trim <- build_scenario(
  base,
  edits    = list(target = list(channel = "TV"), scale = 0.8),
  name     = "Q2 2026 Brand Plan — TV trim",
  nickname = "TV -20%"
)

sum(trim@data$planned_spend)
```

Operations apply in order, so a budget-neutral shift is two of them:

```{r}
shift <- build_scenario(
  base,
  edits = list(
    list(target = list(channel = "TV"),     total = 350000),
    list(target = list(channel = "Social"), total = 200000)
  ),
  name     = "Q2 2026 Brand Plan — reshaped",
  nickname = "TV→Social"
)

sum(shift@data$planned_spend)
```

The four operations, and the distinction that catches people out:

| Operation | Effect on matched rows |
|---|---|
| `set` | each row becomes exactly this |
| `scale` | multiply each row (`1.2` = +20%) |
| `delta` | add to each row (may be negative) |
| `total` | rows **sum** to this, holding their current mix |

`set` is per row; `total` is across rows. "Set the TV budget to 350k" is almost
always `total`.

A target value that does not exist is an error naming the alternatives — never
a silent no-op:

```{r, error = TRUE}
build_scenario(base, edits = list(target = list(channel = "Radio"), scale = 2),
               name = "typo")
```

That matters most when a language model is generating the calls: a quiet no-op
would look exactly like success.

### 2. A data frame — for absolute allocations

Natural for optimizer output, where you already have the numbers:

```{r}
alloc <- data.frame(
  channel       = c("TV", "TV"),
  partner       = c("NBC", "ESPN"),
  week          = c(weeks[1], weeks[1]),
  planned_spend = c(50000, 20000)
)

opt <- build_scenario(base, edits = alloc,
                      name      = "Q2 2026 Brand Plan — optimized",
                      nickname  = "optimized",
                      objective = "max KPI at 70k for w/c Apr 6")
```

Note `objective`: the package never learns *how* the numbers were chosen, so
that free-text field is where you record it.

### 3. A named vector — for one or two cells

Keyed by `line_item()` over the full grain. Terse, and what an editable table
produces:

```{r}
key <- line_item(base@data, base@grain)[1]
key

nudge <- build_scenario(base, edits = stats::setNames(45000, key),
                        name = "Q2 2026 Brand Plan — nudge")
```

## Scenarios and lineage

Deriving never touches the parent, and every scenario records where it came
from:

```{r}
identical(sum(base@data$planned_spend), sum(plan_df$planned_spend))
identical(trim@parent_id, base@id)
```

`status` is deliberately **not** inherited — a scenario forked from an approved
plan is not itself approved:

```{r}
base@status
trim@status
```

## Collecting and comparing

A `ScenarioSet` is the comparison registry. Labels prefer `@nickname`, so a set
under development reads well without renaming the formal plans:

```{r}
set <- scenario_set(base)
set <- add_scenario(set, trim)
set <- add_scenario(set, shift)

set
```

Two levels of comparison:

```{r}
compare_scenarios(set, "summary")
```

```{r}
head(compare_scenarios(set, "cell"), 4)
```

The `"cell"` level compares over the **union** of grain cells across the set, so
a line item one scenario adds and another lacks still gets a real delta rather
than `NA` or a missing row.

## Rolling up

`roll_up()` aggregates to a coarser grain — useful when you need channel totals,
or to find the nearest level at which a response model exists for a brand-new
partner:

```{r}
roll_up(base, c("channel", "week"))@data |> head(4)
```

## Where to next

- `vignette("plan_concepts")` — what a plan *is*, and the boundaries that keep
  it small.
- `?build_scenario` — the full edit-form reference.
- `?check_coverage` — pairing a plan with a decomp.
