---
title: "The plan model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The plan model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(mediaplanr)
```

`mediaplanr` is small on purpose. Most of the decisions that shaped it were
about what to *leave out*, and those decisions are not visible from the function
reference. This vignette covers the model behind the functions — the ideas that,
misunderstood, lead to using the package incorrectly in ways nothing will catch.

`vignette("getting_started")` covers what to type.

## A plan holds intent

Every row of a plan is **planned** spend — what was *intended* — including for
weeks that have already passed. It is never a record of what happened.

That distinction matters because reality diverges from intent constantly:
under-delivery, over-delivery, a partner that never went live. Actualised spend
and attributed results live in a **decomp** — the output of a media-mix model —
which this package does not hold and does not read.

So a row for a past week is still intent. Revising it is an ordinary thing to
do: you are correcting what the plan *said*, not rewriting history.

```{r}
p <- media_plan_from_df(
  data.frame(
    week          = as.Date(c("2026-01-05", "2026-01-05")),
    channel       = c("TV", "Search"),
    planned_spend = c(50000, 20000)
  ),
  grain = c("channel", "week"), week = "week",
  name  = "A plan whose weeks are long past"
)

# Editing a week from January is not special, and nothing objects.
build_scenario(p, edits = list(target = list(channel = "TV"), scale = 1.5),
               name = "revised")@data
```

## A plan is blind to past and future

This follows from the above, and it is the point most likely to be missed.

"Historical", "future", "under-delivery", "over-delivery" are **not properties
of a plan**. They only exist for a *pairing* of a plan with a decomp that
reports actuals through some date. A plan on its own cannot tell you which of
its rows are in the past, because that depends entirely on which decomp you hold
it against — and that boundary moves every time the decomp is refreshed.

Consequences you can see in the API:

- There is no `historical_spend` column. An earlier revision had one; it was
  removed. Freezing a moving boundary into the data guarantees it goes stale.
- No through-date is stored on a `MediaPlan`.
- Nothing stops you editing a row for a past week.

The one pairing operation the package provides is `check_coverage()`:

```{r}
plan <- media_plan_from_df(
  data.frame(
    week          = rep(as.Date(c("2026-03-02", "2026-04-06")), each = 2),
    channel       = c("TV", "Search", "TV", "Search"),
    partner       = c("NBC", "Google", "NBC", "Google"),
    planned_spend = c(30000, 20000, 35000, 25000)
  ),
  grain = c("channel", "partner", "week"), week = "week",
  name  = "Q2 plan"
)

# What the model measured, historically.
decomp <- data.frame(
  channel = c("TV", "Search", "Audio"),
  partner = c("NBC", "Google", "Spotify"),
  stringsAsFactors = FALSE
)

missing <- check_coverage(plan, decomp, through = as.Date("2026-03-15"))
missing
```

Three things about that check are deliberate:

**It runs decomp → plan, not the reverse.** Every line item the model measured
should appear in the plan's pre-through-date rows. The opposite direction is
never flagged: the future portion of a plan is *expected* to introduce new
partners and tactics.

**It is scoped by `through`.** Only rows before that date are considered, since
the decomp only knows about history.

**It warns rather than errors.** Both artifacts are individually valid; it is
their pairing that disagrees. An app should surface that and let the user
resolve it, not refuse to load a perfectly well-formed plan.

That last point is the one exception to the package's general strictness.
Structural problems — negative spend, duplicate rows, a week column that is not
a `Date` — are hard errors, so an invalid plan cannot exist:

```{r, error = TRUE}
media_plan_from_df(
  data.frame(channel = c("TV", "TV"), planned_spend = c(1, 2)),
  grain = "channel", name = "duplicated"
)
```

## Line item, grain, row

Three related ideas that are easy to conflate.

A **line item** is a channel / partner / tactic combination: the *time-free*
identity of something you are buying. It is what a decomp constrains and what
response models attach to.

The **grain** is the set of columns identifying a **row**. At a weekly grain a
row is a line item *for one week*.

```{r}
plan@grain              # what identifies a row
line_item_grain(plan)   # what identifies a line item — the grain minus the week

unique(line_item(plan@data, line_item_grain(plan)))
```

`grain` is deliberately ordered coarsest-first, because `roll_up()` treats it as
a nesting. Dropping the rightmost column gives the next coarsest level — which
is how you find a model for a line item that has none of its own:

```{r}
roll_up(plan, c("channel", "week"))@data
```

The package is a **flat table at a configurable grain**, not a nested object
tree. A plan keyed by `channel`, by `channel + partner`, or by
`channel + partner + tactic + week` is the same class at different grains. A
Channel → Tactic → Flight hierarchy is deferred, not designed out.

## Scenarios are forks

A scenario is a new plan derived from another. The original is never modified —
`MediaPlan` uses value semantics for exactly this reason. If deriving mutated
the parent, the thing you are comparing against would change underneath you.

```{r}
before <- sum(plan@data$planned_spend)
s <- build_scenario(plan, edits = list(scale = 2), name = "double")
after <- sum(plan@data$planned_spend)

c(before = before, after = after, unchanged = identical(before, after))
```

Every derivation records its parent, so lineage chains:

```{r}
s2 <- build_scenario(s, edits = list(scale = 0.5), name = "back down")
identical(s@parent_id, plan@id)
identical(s2@parent_id, s@id)
```

Ids are opaque on purpose. They exist for identity and lineage; human meaning
lives in `@name`, `@nickname` and `@objective`. They are not yet an external
join contract, so do not build one on them.

### Status is not inherited

`@status` moves through a fixed vocabulary:

```{r}
status_levels()
```

A scenario derived from an **approved** plan is not itself approved. Carrying
that forward would manufacture an approval nobody gave, so it resets:

```{r}
approved <- media_plan_from_df(
  data.frame(channel = "TV", planned_spend = 100000),
  grain = "channel", name = "Signed off", status = "approved"
)

child <- build_scenario(approved, edits = list(scale = 1.5), name = "variant")
c(parent = approved@status, child = child@status)
```

`advertiser` and `planner` *are* inherited — they describe the engagement and
the person working, not the individual scenario.

## Comparison unions the cells

`compare_scenarios(level = "cell")` compares over the **union** of grain cells
across the whole set, zero-filling where a scenario has no row.

This is invisible when scenarios come from `build_scenario()`, since those
always inherit the parent's exact row set. It matters when scenarios are
authored independently — one workbook tab each, say — because then their row
sets need not match:

```{r}
a <- media_plan_from_df(
  data.frame(channel = c("TV", "Search"), planned_spend = c(80000, 40000)),
  grain = "channel", name = "Plan", nickname = "baseline"
)
b <- media_plan_from_df(
  data.frame(channel = "TV", planned_spend = 120000),   # Search dropped entirely
  grain = "channel", name = "Plan", nickname = "no search"
)

compare_scenarios(add_scenario(scenario_set(a), b), "cell")
```

Search appears for `no search` as a **−40,000 delta**, not as a missing row.
Without the union, a cut would simply vanish from the table — totals would still
reconcile, and the reader would see money appearing from nowhere.

## What lives elsewhere

`mediaplanr` does not fit models, forecast, or optimize, and imports nothing but
S7.

```
         mediaplanr                       mrmopt
  ┌────────────────────────┐   ┌──────────────────────────┐
  │ plan object, grain,    │   │ response curves, fitting,│
  │ validation, lineage,   │   │ forecasting, budget      │
  │ scenarios, comparison  │   │ optimization (opt_mix)   │
  └────────────────────────┘   └──────────────────────────┘
                ↘                   ↙
                  a caller wires them
```

An earlier revision *did* include a `forecast()` and a budget optimizer. Both
were removed, for reasons worth stating so they are not rebuilt by accident:

- The consuming application already called `mrmopt::opt_mix()`, so the
  in-package optimizer was a weaker second implementation.
- Implementing `forecast()` forced a **modelling** decision into a container:
  response curves have a non-zero lower asymptote, so "predicted volume" is
  ambiguous, and the package had to pick a convention. That belongs where the
  modelling is owned.
- It inverted the dependency. A `ScenarioSet` could not exist without a set of
  fitted models, so plans could not be held or compared without them.

An allocation computed by an optimizer re-enters through the same door a hand
edit uses:

```{r, eval = FALSE}
result <- mrmopt::opt_mix(models, budget = 160000)
alloc  <- data.frame(channel = result$solution$channel,
                     planned_spend = result$solution$optimal_spend)

build_scenario(base, edits = alloc, name = "Optimized",
               objective = "max KPI at 160k, opt_mix point method")
```

The package never learns where the numbers came from. `@objective` is where you
record why.
