---
title: "The plan model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The plan model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(mediaplanr)
```

`mediaplanr` is small on purpose. Most of the decisions that shaped it were
about what to *leave out*, and those decisions are not visible from the function
reference. This vignette covers the model behind the functions — the ideas that,
misunderstood, lead to using the package incorrectly in ways nothing will catch.

`vignette("getting_started")` covers what to type.

## A plan holds intent

Every row of a plan is **planned** spend — what was *intended* — including for
weeks that have already passed. It is never a record of what happened.

That distinction matters because reality diverges from intent constantly:
under-delivery, over-delivery, a partner that never went live. Actualised spend
and attributed results live in a **decomp** — the output of a media-mix model —
which this package does not hold and does not read.

So a row for a past week is still intent. Revising it is an ordinary thing to
do: you are correcting what the plan *said*, not rewriting history.

```{r}
p <- media_plan_from_df(
  data.frame(
    week          = as.Date(c("2026-01-05", "2026-01-05")),
    channel       = c("TV", "Search"),
    planned_spend = c(50000, 20000)
  ),
  grain = c("channel", "week"), week = "week",
  name  = "A plan whose weeks are long past"
)

# Editing a week from January is not special, and nothing objects.
build_scenario(p, edits = list(target = list(channel = "TV"), scale = 1.5),
               name = "revised")@data
```

## A plan is blind to past and future

This follows from the above, and it is the point most likely to be missed.

"Historical", "future", "under-delivery", "over-delivery" are **not properties
of a plan**. They only exist for a *pairing* of a plan with a decomp that
reports actuals through some date. A plan on its own cannot tell you which of
its rows are in the past, because that depends entirely on which decomp you hold
it against — and that boundary moves every time the decomp is refreshed.

Consequences you can see in the API:

- There is no `historical_spend` column. An earlier revision had one; it was
  removed. Freezing a moving boundary into the data guarantees it goes stale.
- No through-date is stored on a `MediaPlan`.
- Nothing stops you editing a row for a past week.

The one pairing operation the package provides is `check_coverage()`:

```{r}
plan <- media_plan_from_df(
  data.frame(
    week          = rep(as.Date(c("2026-03-02", "2026-04-06")), each = 2),
    channel       = c("TV", "Search", "TV", "Search"),
    partner       = c("NBC", "Google", "NBC", "Google"),
    planned_spend = c(30000, 20000, 35000, 25000)
  ),
  grain = c("channel", "partner", "week"), week = "week",
  name  = "Q2 plan"
)

# What the model measured, historically.
decomp <- data.frame(
  channel = c("TV", "Search", "Audio"),
  partner = c("NBC", "Google", "Spotify"),
  stringsAsFactors = FALSE
)

missing <- check_coverage(plan, decomp, through = as.Date("2026-03-15"))
missing
```

Three things about that check are deliberate:

**It runs decomp → plan, not the reverse.** Every line item the model measured
should appear in the plan's pre-through-date rows. The opposite direction is
never flagged: the future portion of a plan is *expected* to introduce new
partners and tactics.

**It is scoped by `through`, on overlap.** A row is considered when **any part
of its in-market period falls before** that date, since the decomp only knows
about history. A decomp reporting through a Wednesday has measured part of the
week that began on the Monday, so that line item should be expected in the plan.

The comparison is against the row's real period, not the bare week value, so it
follows a flight that starts mid-week:

```{r}
midweek <- media_plan_from_flights(
  data.frame(channel = "TV", flight_start = as.Date("2026-04-08"),
             flight_end = as.Date("2026-04-19"), planned_spend = 100),
  grain = "channel", name = "Starts on the Wednesday"
)

# the buy was not live on the 7th, so nothing is in scope to check
tryCatch(check_coverage(midweek, data.frame(channel = "TV"),
                        through = as.Date("2026-04-07")),
         warning = function(w) conditionMessage(w))
```

Being in scope says the decomp saw *part* of a buy, not much of it: a long
flight that began the day before `through` qualifies on a single measured day.
That is the right answer for the question being asked — should this line item
appear in the plan? — but it is not a statement about delivery.

**It warns rather than errors.** Both artifacts are individually valid; it is
their pairing that disagrees. An app should surface that and let the user
resolve it, not refuse to load a perfectly well-formed plan.

### Extent is not a boundary

A plan does know one thing about its own dates: **which ones it covers**. That
is its *extent*, and it is not the same species of fact as a through-date.

```{r}
c(plan@flight_start, plan@flight_end)
plan@flight_days
```

The discriminating question is: **can this value change without the plan
changing?** A decomp's through-date can — it moves every refresh, and the plan
sits still while it does. A flight window cannot: it is a pure function of
`@data`. So one is an argument to `check_coverage()` and the other is a property
of the plan.

That test is also what keeps the class from re-growing the things that were
removed. `historical_spend`, `weeks_remaining`, `is_active`, `pct_delivered` all
fail it — each needs either `Sys.Date()` or a decomp — so none of them belongs
here, however convenient they would be.

Note that `@flight_end` is the last day the plan is in market, not the last week
*start* stored in `@data`. A week is seven days of spend, so a plan whose final
week begins on a Monday runs through the Sunday after it.

That last point about warnings is the one exception to the package's general
strictness.
Structural problems — negative spend, duplicate rows, a week column that is not
a `Date` — are hard errors, so an invalid plan cannot exist:

```{r, error = TRUE}
media_plan_from_df(
  data.frame(channel = c("TV", "TV"), planned_spend = c(1, 2)),
  grain = "channel", name = "duplicated"
)
```

## Line item, grain, row

Three related ideas that are easy to conflate.

A **line item** is a channel / partner / tactic combination: the *time-free*
identity of something you are buying. It is what a decomp constrains and what
response models attach to.

The **grain** is the set of columns identifying a **row**. At a weekly grain a
row is a line item *for one week*.

```{r}
plan@grain              # what identifies a row
line_item_grain(plan)   # what identifies a line item — the grain minus the week

unique(line_item(plan@data, line_item_grain(plan)))
```

`line_item()` gives you the key for every *row*. For the distinct roster — one
row per line item, with its total, its share of the plan's rows and its own
flight window — use `line_item_summary()`:

```{r}
line_item_summary(plan)
```

The grain values come back as columns rather than folded into the key, because
that is the shape an edit `target` takes. So "which channels are in this plan?"
is answered with values you can hand straight to `build_scenario()`:

```{r}
grain_values(plan, "channel")
```

There is deliberately no `channels()`. The package never privileges a column
name — `channel` is simply a dimension you chose to key on, and another plan may
key on `media_type` or `vehicle`. Naming your own column keeps that true, and a
typo errors with the alternatives listed rather than returning nothing.

`grain` is deliberately ordered coarsest-first, because `roll_up()` treats it as
a nesting. Dropping the rightmost column gives the next coarsest level — which
is how you find a model for a line item that has none of its own:

```{r}
roll_up(plan, c("channel", "week"))@data
```

The package is a **flat table at a configurable grain**, not a nested object
tree. A plan keyed by `channel`, by `channel + partner`, or by
`channel + partner + tactic + week` is the same class at different grains.
Nested plans — a channel team's detail rolling up into the weekly topline — are
deferred, not designed out; see `Roadmap.md`.

## Flights are authored, weeks are held

A planner states a buy as a **flight**: in-market dates and a total. A plan is
*held*, charted and edited as weeks. These are different things, and the package
keeps them apart rather than picking one.

`media_plan_from_flights()` expands a flight onto the weekly rows `@data`
already holds, so everything downstream sees the ordinary shape. `flights()`
goes back the other way — and it is **exact, never inferred**:

```{r}
weekly <- media_plan_from_df(
  data.frame(week = seq(as.Date("2026-04-06"), by = "week", length.out = 4),
             channel = "TV", planned_spend = rep(30000, 4)),
  grain = c("channel", "week"), week = "week", name = "Four equal weeks"
)

nrow(flights(weekly))
```

Nothing is reported, because nothing was authored. Four consecutive 30,000 weeks
are equally consistent with one 28-day flight, four weekly buys, or two
fortnightly ones, and no rule can tell them apart. Guessing from runs of equal
spend would be a guess presented as a fact — the failure mode the package
refuses everywhere, the same reason a mistyped target errors instead of
quietly matching nothing.

So the inverse works only because each buy's identity is *stored* when it is
authored. That identity survives derivation, which is what lets
`compare_scenarios(level = "flight")` say a buy **moved** rather than showing an
unexplained subtraction in one week and an addition in another.

### One period per line item per week

A row is still one line item for one period, so two separate buys of the same
line item cannot overlap the same week. That is an error naming the fix — add a
column that tells them apart — rather than a silent sum that would merge two
distinct buys.

It is also the price of keeping `@data` in the long weekly shape that makes the
package pleasant to chart, edit and export. Making the flight the stored row
instead was tried and rejected: `@grain`'s time member would become the flight
start, so two scenarios that stage the same money differently would share no
cells, and every reshaping would read as an add plus a drop rather than a move.

### Pacing records shape, not history

A flight has a **shape** as well as a size. By default its spend is spread
evenly across its days, so its weekly figures are whatever that gives — which
for a buy with ragged ends is not four equal numbers:

```{r}
ragged <- media_plan_from_flights(
  data.frame(channel = "OOH", flight_start = as.Date("2026-04-08"),
             flight_end = as.Date("2026-05-03"), planned_spend = 100000),
  grain = "channel", name = "Starts on a Wednesday"
)

ragged@data$planned_spend      # 5 days, then 7, 7, 7
flights(ragged)$pacing
```

`pacing` is **re-derived after every edit**, not remembered. The question it
answers is "do these rows still match an even spread?", never "did somebody
touch this?" — and that distinction is what makes it behave sensibly:

```{r}
shape <- function(p) flights(p)$pacing

# scaling changes the buy's SIZE, not its shape
shape(build_scenario(ragged, edits = list(scale = 0.5), name = "half"))

# a flat amount on every week cannot stay proportional to days, so it does
shape(build_scenario(ragged, edits = list(delta_each = 1000), name = "flat"))
```

Because it is derived rather than remembered, shaping a flight by hand and then
shaping it back reports `"even"` again. There is no permanent mark.

A hand-shaped flight is never **broken** by the edit: it keeps its identity, so
`flights()` still reports it whole and it can still be compared, moved and
restaged as one buy. `"custom"` says the shape is no longer a rule, not that the
buy has stopped being a buy.

### The one place the package rewrites a number

Worth knowing because it is the exception to everything else here. When a flight
is still evenly paced, its rows are **snapped** to the canonical allocation —
the same figures a fresh import would produce.

The reason is drift. Scaling values already rounded to the cent does not land on
the cent-exact allocation of the scaled total, so without snapping a long chain
of scenarios would accumulate fractions of a cent and slowly stop being even at
all. With it, the chain is stable:

```{r}
chain <- ragged
for (i in 1:12) {
  chain <- build_scenario(chain, edits = list(scale = 1.37), name = paste0("s", i))
}
back <- build_scenario(chain, edits = list(total = 100000), name = "back to 100k")

identical(back@data$planned_spend, ragged@data$planned_spend)
```

The snap only touches flights already even to within a rounding tolerance, moves
at most a cent between weeks, and **never changes a flight's total**. A
hand-shaped flight is left exactly as the planner set it.

## The rate is what survives an edit

Spend is the **common currency** — the one quantity every channel shares — so it
stays mandatory. A line item may also record what it buys, how much, and at what
price, bound by one identity:

```
planned_spend = planned_units * planned_rate / rate_per(unit_type)
```

Supply any two and the third is computed, whichever way you actually have it.
All three are then stored and the validator enforces the identity, so they
cannot drift apart.

The question that shapes everything else is: **when spend changes, which of the
other two gives way?** A negotiated CPM does not improve because the budget was
cut. So the rate holds and the units follow:

```{r}
social <- media_plan_from_df(
  data.frame(channel = "Social", unit_type = "impression",
             planned_spend = 140000, planned_rate = 5),
  grain = "channel", name = "Social"
)

cut <- build_scenario(social, edits = list(delta = -40000), name = "trimmed")
cut@data[, c("planned_spend", "planned_units", "planned_rate")]
```

That is one rule, applied wherever money moves — through scaling, through a
flight spreading across its weeks, through `calendarize()`. Had units been the
invariant instead, every one of those places would have needed its own rule for
keeping units in step.

Units add up only **within** one unit type. Across a mix, `roll_up()` keeps the
spend and reports no units, because a total that adds GRPs to clicks is a number
nobody should act on — and the spend is exactly what a common currency is for.

## Scenarios are forks

A scenario is a new plan derived from another. The original is never modified —
`MediaPlan` uses value semantics for exactly this reason. If deriving mutated
the parent, the thing you are comparing against would change underneath you.

```{r}
before <- sum(plan@data$planned_spend)
s <- build_scenario(plan, edits = list(scale = 2), name = "double")
after <- sum(plan@data$planned_spend)

c(before = before, after = after, unchanged = identical(before, after))
```

Every derivation records its parent, so lineage chains:

```{r}
s2 <- build_scenario(s, edits = list(scale = 0.5), name = "back down")
identical(s@parent_id, plan@id)
identical(s2@parent_id, s@id)
```

Ids are opaque on purpose. They exist for identity and lineage; human meaning
lives in `@name`, `@nickname` and `@objective`. They are not yet an external
join contract, so do not build one on them.

### Status is not inherited

`@status` moves through a fixed vocabulary:

```{r}
status_levels()
```

A scenario derived from an **approved** plan is not itself approved. Carrying
that forward would manufacture an approval nobody gave, so it resets:

```{r}
approved <- media_plan_from_df(
  data.frame(channel = "TV", planned_spend = 100000),
  grain = "channel", name = "Signed off", status = "approved"
)

child <- build_scenario(approved, edits = list(scale = 1.5), name = "variant")
c(parent = approved@status, child = child@status)
```

`advertiser` and `planner` *are* inherited — they describe the engagement and
the person working, not the individual scenario.

## Comparison unions the cells

`compare_scenarios(level = "cell")` compares over the **union** of grain cells
across the whole set, zero-filling where a scenario has no row.

This is invisible when scenarios come from `build_scenario()`, since those
always inherit the parent's exact row set. It matters when scenarios are
authored independently — one workbook tab each, say — because then their row
sets need not match:

```{r}
a <- media_plan_from_df(
  data.frame(channel = c("TV", "Search"), planned_spend = c(80000, 40000)),
  grain = "channel", name = "Plan", nickname = "baseline"
)
b <- media_plan_from_df(
  data.frame(channel = "TV", planned_spend = 120000),   # Search dropped entirely
  grain = "channel", name = "Plan", nickname = "no search"
)

compare_scenarios(add_scenario(scenario_set(a), b), "cell")
```

Search appears for `no search` as a **−40,000 delta**, not as a missing row.
Without the union, a cut would simply vanish from the table — totals would still
reconcile, and the reader would see money appearing from nowhere.

## What lives elsewhere

`mediaplanr` does not fit models, forecast, or optimize, and imports nothing but
S7.

```
         mediaplanr                       mrmopt
  ┌────────────────────────┐   ┌──────────────────────────┐
  │ plan object, grain,    │   │ response curves, fitting,│
  │ validation, lineage,   │   │ forecasting, budget      │
  │ scenarios, comparison  │   │ optimization (opt_mix)   │
  └────────────────────────┘   └──────────────────────────┘
                ↘                   ↙
                  a caller wires them
```

An earlier revision *did* include a `forecast()` and a budget optimizer. Both
were removed, for reasons worth stating so they are not rebuilt by accident:

- The consuming application already called `mrmopt::opt_mix()`, so the
  in-package optimizer was a weaker second implementation.
- Implementing `forecast()` forced a **modelling** decision into a container:
  response curves have a non-zero lower asymptote, so "predicted volume" is
  ambiguous, and the package had to pick a convention. That belongs where the
  modelling is owned.
- It inverted the dependency. A `ScenarioSet` could not exist without a set of
  fitted models, so plans could not be held or compared without them.

An allocation computed by an optimizer re-enters through the same door a hand
edit uses:

```{r, eval = FALSE}
result <- mrmopt::opt_mix(models, budget = 160000)
alloc  <- data.frame(channel = result$solution$channel,
                     planned_spend = result$solution$optimal_spend)

build_scenario(base, edits = alloc, name = "Optimized",
               objective = "max KPI at 160k, opt_mix point method")
```

The package never learns where the numbers came from. `@objective` is where you
record why.
